---
title: "Introduction to hiaR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to hiaR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This example shows how to use the `hiaR` package on a realistic dataset. The `hiaR`
package is a wrapper for a Java application which implements the DYNAMO-HIA model.
(TODO: Add model explanation).

We start by loading the `hiar` package:

```{r setup}
library(hiaR)
```

We apply the model to a dataset from the Netherlands. The model requires reference data. 
The reference data for the Netherlands is included in the package. We move the reference data into
a temporary local directory with:

```{r}
# Create local temporary directory; the directory will be deleted when the R session is finished
temp_dir <- withr::local_tempdir()

fs::dir_copy(system.file("extdata", "example-nl", package = "hiaR"), temp_dir, overwrite = TRUE)
```

The `Reference_Data` folder in our directory now contains the reference data for 
the Netherlands. The reference data contains information about the Dutch population, diseases, and risk factors. These will be used in the simulations. We can get an overview of the available populations, diseases, and risk factors with helper functions from the `hiaR` package:

```{r}
# Define path to reference data
reference_data_path <- fs::path(temp_dir, "Reference_Data")

cat("Available diseases:\n")
disease_names <- get_disease_names(reference_data_path)

cat("\nAvailable populations:\n")
population_names <- get_population_names(reference_data_path)

cat("\nAvailable risk factors:\n")
risk_factor_names <- get_risk_factor_names(reference_data_path)
```

We will now configure a simulation to assess the effect of the risk factor smoking on the health of the Dutch population. We will use a risk factor configuration with three categories (never smoked, former smoker, current smoker). To configure the risk factor, we need to provide a unique name, the name of the file with the risk factor transition type, and the name of the file with the risk factor prevalence. We can get the names of the files from the reference data with the `get_risk_factor_filenames` function:


```{r}
get_risk_factor_filenames(reference_data_path, "Smoking_cat3")
```
The files in the different sub directories are briefly described here (for details the DYNAMO-HIA documentation):

- `Prevalence`: Files with data on which part of the population is in which risk factor category (split by age and gender)
- `Relative_Risks_For_Death`: Files with the relative risks for death for each risk factor category (split by age and gender)
- `Relative_Risks_For_Disability` Files with the relative risks for disability for each risk factor category (split by age and gender)
- `Transitions`: Files with the transition probabilities between the risk factor categories

With the file names we can configure the risk factor:

```{r}
smoking_config <- configure_risk_factor(
  name = "Smoking_cat3", # Should be the same as the name in the reference data
  transition_filename = "NL_RF_smoking_Transitions_Netto",
  prevalence_filename = "NL_RF_smoking_cat3_Prev_V1"
)
```

We chose a "Netto" transition file which indicates that the risk factor prevalence for each age and gender group will remain the same during the simulation.

The DYNAMO-HIA model also takes different diseases within the population into account. While the reference data provides information about several diseases we will only look at lung cancer in this example. The disease configuration requires a unique name for the disease, names of files with the disease prevalence, incidence, excess mortality, and disability weights. We can get the names of the files from the reference data with the `get_disease_filenames` function:

```{r}
get_disease_filenames(reference_data_path, "Lung_Cancer")
```

A brief description of the sub directories:

- `Disability`: Files with the disability (disability-adjusted life years) weights for the disease (split by age and gender)
- `Excess_Mortalities`: Files with the additional mortality when carrying the disease (split by age and gender)
- `Incidence`: Files with the incidence of the disease (split by age and gender)
- `Prevalence`: Files with the prevalence of the disease (split by age and gender)
- `Relative_Risks_From_Diseases`: Files with the relative risks of carrying the disease when having another disease (split by age and gender; not used in this example)
- `Relative_Risks_From_Risk_Factor`: Files with the relative risks of carrying the disease when being exposed to a risk factor (split by age and gender)

With the file names we can configure the disease:

```{r}
lung_cancer_config <- configure_disease(
  name = "Lung_Cancer",
  prevalence_filename = "NL_disease_LungCa_Prev_V2",
  incidence_filename = "NL_disease_LungCa_Inc_V2",
  excess_mortality_filename = "NL_disease_LungCa_ExecMor_V2",
  disability_weights_filename = "NL_disease_LungCa_DALY_V1"
)
```

We can also configure a relative risk of having lung cancer due to smoking. The relative risk configuration requires an index, the name of the risk factor, the name of the disease, and the name of the file with the relative risk data:

```{r}
lung_cancer_smoking_rr_config <- configure_relative_risk(
  index = 0, # Start with zero
  from = "Smoking_cat3",
  to = "Lung_Cancer",
  relative_risk_filename = "RR_to_LungCa-Smoking_cat3"
)
```

Besides risk factors and diseases, we also need to configure the scenario that we want to simulate. Scenarios require a unique name, a success rate for an intervention, the minimum and maximum age as well as the targeted gender for the intervention. Most importantly, they need names of files with the transition type and prevalence of the risk factor which is changed due to the intervention.

We create a scenario that intervenes on the risk factor smoking and assumes that the targeted population never smoked. We choose this intervention by specifying a prevalence file that only contains never smokers.

```{r}
smoking_intervention = configure_scenario(
  name = "All_Never_Smokers",
  success_rate = 80,
  min_age = 18,
  max_age = 65,
  gender = 2, # Both female and male
  transition_filename = "NL_RF_smoking_Transitions_Netto", # Same as in reference scenario
  prevalence_filename = "RF_smoking_cat_All_Never_Smokers_Prev" # Different from reference scenario
)
```

Now that we have configured a risk factor, disease, and scenario, we write them to an XML file in a new subfolder in the `Simulations` folder using the `create_simulation_dir` function. The function requires the name of the simulation and additional parameters to configure the simulation. We will simulate the effect of the smoking intervention on the Dutch population for 10 years starting in 2010. We will use a random seed of 42 to ensure reproducibility.

```{r}
simulation_name <- "Netherlands_Smoking_Intervention"

create_simulation_dir(
  simulation_name = fs::path(temp_dir, "Simulations", simulation_name),
  population_name = "Netherlands",
  has_newborns = TRUE,
  starting_year = 2010,
  number_of_years = 10,
  population_size = 10,
  min_age = 0,
  max_age = 95,
  time_step = 1,
  ref_scenario_name = "Netherlands_Reference_Scenario",
  random_seed = 42,
  result_type = "Emptyness2BFilled",
  scenarios = list(smoking_intervention), # Wrap in list because we could have multiple scenarios
  diseases = list(lung_cancer_config), # Wrap in list because we could have multiple diseases
  risk_factors = smoking_config, # No list because we only have one risk factor
  relative_risks = list(lung_cancer_smoking_rr_config) # Wrap in list because we could have multiple relative risks
)
```
The function returns `TRUE` indicating that our configurations are valid and the subfolder was successfully created. We can look at all the available simulations with the `get_simulation_names` function:

```{r}
get_simulation_names(temp_dir)
```

The subfolder will also contain the results from the simulation.

Before we can run the simulation, we need to add the simulation name to a batch file which we will use as input for the DYNAMO-HIA Java application.

```{r}
batch_file_name <- "simulation_names.txt"

write(simulation_name, fs::path(temp_dir, batch_file_name))
```

We can now run the simulation using the `run_dynamo_hia` function. The function takes the path to the simulation folder as an argument.

```{r, eval=FALSE}
# TODO: Add dynamo path and download
withr::with_dir(temp_dir, {
  run_dynamo_hia(batch_file_name, Sys.getenv("DYNAMO_HIA_PATH"))
})
```
The function returns `TRUE` indicating that the simulation was successful. The results of the simulation are stored in the folder of the simulation.

(TODO: Explanation of the results and how to visualize them).
