---
title: "Introduction to hiaR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to hiaR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.dim = c(7, 5)
)
```

This introduction shows how to use the `hiaR` package in a simple example. The `hiaR`
package is a wrapper for a Java application which implements the DYNAMO-HIA model. 
For details on the DYNAMO-HIA model, see the [DYNAMO-HIA documentation](https://dynamo-hia.github.io/). 
The DYNAMO-HIA model simulates the health impact of interventions on risk factors on 
disease outcomes and life expectancy in a population. In this example, we simulate the effect of an
intervention on smoking on the development of lung cancer in the Dutch population.

We start by loading the `hiar` package:

```{r setup}
library(hiaR)
```

## Reference data

To run simulations, the DYNAMO-HIA model requires reference data. The reference data for 
the Netherlands is included in the package. We move the reference data into
a temporary local directory using the [`withr`](https://withr.r-lib.org/) package:

```{r}
# Create local temporary directory; the directory will be deleted when the R session is finished
temp_dir <- withr::local_tempdir()

fs::dir_copy(system.file("extdata", "example-nl", package = "hiaR"), temp_dir, overwrite = TRUE)
```

The `Reference_Data` folder in our directory now contains the reference data for 
the Netherlands including information about the Dutch population, diseases, and risk factors. 
These will be used mainly as the *reference scenario* in the simulations. 
We can get an overview of the available populations, 
diseases, and risk factors with helper functions from the `hiaR` package:

```{r}
# Define path to reference data
reference_data_path <- fs::path(temp_dir, "Reference_Data")

cat("Available diseases:\n")
disease_names <- get_disease_names(reference_data_path)

cat("\nAvailable populations:\n")
population_names <- get_population_names(reference_data_path)

cat("\nAvailable risk factors:\n")
risk_factor_names <- get_risk_factor_names(reference_data_path)
```

## Configuration

We will now configure a simulation to assess the effect of the risk factor smoking on the health 
of the Dutch population. We will use a risk factor configuration with three categories 
(never smoked, former smoker, current smoker). To configure the risk factor, we need to provide a 
unique name, the name of the file with the risk factor transition type, and the name of the file 
with the risk factor prevalence. We can get the names of the files from the reference data with 
the `get_risk_factor_filenames` function:


```{r}
get_risk_factor_filenames(reference_data_path, "Smoking_cat3")
```
The files in the different sub directories are briefly described here 
(for details, see the [DYNAMO-HIA documentation](https://dynamo-hia.github.io/)):

- `Prevalence`: Files with data on which part of the population is in which risk factor category 
(split by age and gender)
- `Relative_Risks_For_Death`: Files with the relative risks for death for each risk factor category 
(split by age and gender)
- `Relative_Risks_For_Disability` Files with the relative risks for disability for each risk factor 
category (split by age and gender)
- `Transitions`: Files with the transition probabilities between the risk factor categories

With the file names we can configure the risk factor:

```{r}
smoking_config <- configure_risk_factor(
  name = "Smoking_cat3", # Should be the same as the name in the reference data
  transition_filename = "NL_RF_smoking_Transitions_Netto",
  prevalence_filename = "NL_RF_smoking_cat3_Prev_V1"
)
```

We chose a "Netto" transition file which indicates that the risk factor prevalence for each age and 
gender group will remain the same during the simulation.

The DYNAMO-HIA model also takes different diseases within the population into account. While the 
reference data provides information about several diseases we will only look at lung cancer in 
this example. The disease configuration requires a unique name for the disease, names of files with
the disease prevalence, incidence, excess mortality, and disability weights. We can get the names 
of the files from the reference data with the `get_disease_filenames` function:

```{r}
get_disease_filenames(reference_data_path, "Lung_Cancer")
```

A brief description of the sub directories:

- `Disability`: Files with the disability (disability-adjusted life years) weights for the disease 
(split by age and gender)
- `Excess_Mortalities`: Files with the additional mortality when carrying the disease 
(split by age and gender)
- `Incidence`: Files with the incidence of the disease (split by age and gender)
- `Prevalence`: Files with the prevalence of the disease (split by age and gender)
- `Relative_Risks_From_Diseases`: Files with the relative risks of carrying the disease when having 
another disease (split by age and gender; not used in this example)
- `Relative_Risks_From_Risk_Factor`: Files with the relative risks of carrying the disease when 
being exposed to a risk factor (split by age and gender)

With the file names we can configure the disease:

```{r}
lung_cancer_config <- configure_disease(
  name = "Lung_Cancer",
  prevalence_filename = "NL_disease_LungCa_Prev_V2",
  incidence_filename = "NL_disease_LungCa_Inc_V2",
  excess_mortality_filename = "NL_disease_LungCa_ExecMor_V2",
  disability_weights_filename = "NL_disease_LungCa_DALY_V1"
)
```

We can also configure a relative risk of having lung cancer due to smoking. The relative risk 
configuration requires an index, the name of the risk factor, the name of the disease, and the name
of the file with the relative risk data:

```{r}
lung_cancer_smoking_rr_config <- configure_relative_risk(
  index = 0, # Start with zero
  from = "Smoking_cat3",
  to = "Lung_Cancer",
  relative_risk_filename = "RR_to_LungCa-Smoking_cat3"
)
```

Besides risk factors and diseases, we also need to configure the intervention scenario that we want
to simulate. Scenarios require a unique name, a success rate for an intervention, the minimum and 
maximum age as well as the targeted gender for the intervention. Most importantly, they need names 
of files with the transition type and prevalence of the risk factor which is changed due to the 
intervention.

We create a scenario that intervenes on the risk factor smoking and assumes that the targeted 
population never smoked. We choose this intervention by specifying a prevalence file that only 
contains "never smokers".

```{r}
smoking_intervention <- configure_scenario(
  name = "All_Never_Smokers",
  transition_filename = "NL_RF_smoking_Transitions_Netto", # Same as in reference scenario
  prevalence_filename = "RF_smoking_cat_All_Never_Smokers_Prev" # Different from reference scenario
)
```

Now that we have configured a risk factor, disease, and scenario, we write them to an XML file in 
a new subfolder in the `Simulations` folder using the `create_simulation_dir` function. 
The function requires the name of the simulation and additional parameters to configure the 
simulation. We will simulate the effect of the smoking intervention on the Dutch population 
for 10 years starting in 2010. We will use a random seed of 42 to ensure reproducibility.

```{r}
simulation_name <- "Netherlands_Smoking_Intervention"

create_simulation_dir(
  simulation_name = fs::path(temp_dir, "Simulations", simulation_name),
  population_name = "Netherlands",
  has_newborns = TRUE,
  starting_year = 2010,
  number_of_years = 10,
  population_size = 10,
  min_age = 0,
  max_age = 95,
  time_step = 1,
  ref_scenario_name = "Netherlands_Reference_Scenario",
  random_seed = 42,
  scenarios = list(smoking_intervention), # Wrap in list because we could have multiple scenarios
  diseases = list(lung_cancer_config), # Wrap in list because we could have multiple diseases
  risk_factors = smoking_config, # No list because we only have one risk factor
  relative_risks = list(lung_cancer_smoking_rr_config) # Wrap in list because we could have multiple relative risks
)
```
The function returns `TRUE` indicating that our configurations are valid and the subfolder was 
successfully created. We can look at all the available simulations with 
the `get_simulation_names` function:

```{r}
get_simulation_names(temp_dir)
```

The subfolder will also contain the results from the simulation.

We need to add the simulation name to a batch file which we 
will use as input for the DYNAMO-HIA Java application.

```{r}
batch_file_name <- "simulation_names.txt"

write(simulation_name, fs::path(temp_dir, batch_file_name))
```

## Running DYNAMO-HIA

Before we can run the simulation, we need to download the DYNAMO-HIA Java application. We can do
this automatically with the `download_github_release` function. The function will automatically
download the correct version for our operating system and place it into our temporary directory.

```{r}
download_github_release(
  repo_url = "https://github.com/Dynamo-HIA/dynamo-hia",
  release_tag = "v3.0.0-beta.1",
  dest_dir = temp_dir
)
```

We can now run the simulation using the `run_dynamo_hia` function. The function takes the path 
to the batch file and the path to the DYNAMO-HIA executable as arguments. The latter is set 
automatically by the function `download_github_release` and stored in the 
environment variable `DYNAMO_HIA_PATH`.

```{r}
withr::with_dir(temp_dir, {
  run_dynamo_hia(batch_file_name, fs::path(temp_dir, Sys.getenv("DYNAMO_HIA_PATH")))
})
```

The function returns `TRUE` indicating that the simulation was successful. The results of the 
simulation are stored in the simulation folder.

## Loading and visualizing results

We can load the results with the `load_simulation_results` function. The function requires the name
of the simulation and the path to the working directory. We start by loading the population results
(using `type = batchoutput`):

```{r}
df_population <- load_simulation_results(temp_dir, simulation_name, type = "batchoutput")

# Get an overview of the data frame
head(df_population)

```

With the population results, we can plot the population pyramid for the starting year of our 
simulation showing the difference between our smoking intervention scenario and the 
reference scenario:

```{r}
plot_population_pyramid(df_population, year = 2010, scenario = 1, difference = TRUE)

```

We can see that disability is more prevalent in older age groups and lung cancer only occurs in 
age groups above 50 and is generally a rare disease.

We can also plot the prevalence of lung cancer at a specific age (75) over time:

```{r}
plot_prevalence(df_population, disease = "Lung_Cancer", age = 75)

```

The plot shows that the prevalence of lung cancer decreases over the years in our intervention 
scenario.

We can also compare the mortality across scenarios over time:

```{r}
plot_mortality(df_population, age = 75)

```

It seems like our intervention has a positive effect on mortality rates in this specific age group.

Finally, we can plot the simulated life expectancy. This requires the life expectancy results which
we can load using `type = cohortLE`:

```{r}
df_life_expectancy <- load_simulation_results(temp_dir, simulation_name, type = "cohortLE")

# Get an overview of the data frame
head(df_life_expectancy)

```

We can now plot the life expecaancy for the different scenarios and genders at a specific age group:

```{r}
plot_life_expectancy(df_life_expectancy, age = 75)

```

We can see that our intervention increases life expectancy in this age group.

## Conclusion

In this example, we showed how to use the `hiaR` package to configure and run a simulation of the
DYNAMO-HIA model using the effect of smoking on lung cancer in the Dutch population as an example.
