---
title: "Creating reference data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating reference data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


This tutorial shows how to use the `hiaR` package to create reference data that can be used in the
DYNAMO-HIA model. We use the Global Burden of Disease (GBD) data to create reference data for
the mortality of COVID-19 in the Netherlands. The GBD data is available from the [IHME website](http://ghdx.healthdata.org/gbd-results-tool). 
The data can be downloaded from the following
[link](https://vizhub.healthdata.org/gbd-results?params=gbd-api-2021-permalink/e97296bbd66d629c3a05e9b309867160)
after registering on the website and entering a user agreement. For simplicity, we include simulated
data with identical format in the `hiaR` package that we will use in this tutorial. Please
install the `dplyr` and `tidyr` packages to run this tutorial:

```{r}
install.packages(c("dplyr", "tidyr"))
```

We start by loading the `hiaR` package and the example data `covid19nl`. When using the real GBD data,
the data can be loaded from CSV with `read.csv` or a similar function. For documentation on the 
variables, see the [IHME Data and Tools Guide](https://www.healthdata.org/research-analysis/about-gbd/gbd-data-and-tools-guide).

```{r setup}
library(dplyr)
library(tidyr)

library(hiaR)

disease <- "COVID-19"

data(covid19nl)

# Take a glimpse at the data
head(covid19nl)
```

The DYNAMO-HIA model requires reference data to be in a specific format. We therefore need to 
reshape and preprocess the raw GBD data. We start by creating data for age groups 0 to 95 years 
that are required by DYNAMO-HIA (under default settings). The GBD data contains 5-year age groups, 
so we need to interpolate these to 1-year age groups by imputing the 5-year rows for each 1-year 
group. We first look at the age groups present in the GBD data.

```{r}
unique_age_groups <- unique(covid19nl$Age)

print(unique_age_groups)

```

We see 19 5-year age groups and one "95+ years" age group that we will treat as the 95-year age group.
We extract the range of each group from the character string and create additional rows for each 1-year age group.
We also recode character values in the `Sex` column as numeric values.

```{r}
preprocessed_df <- covid19nl |>
  mutate(Age = ifelse(Age == "< 5 years",
                      "0-4 years",
                      ifelse(Age == "95+ years",
                             "95 years",
                             Age)),
         sex_num = ifelse(Sex == "Males", 0, 1),
         age_num = gregexpr("\\d+", Age) |> 
           regmatches(Age, m = _) |> 
           sapply(function(x) list(c(x[1]:x[length(x)])))) |>
  unnest(age_num)

```

Now, we can create a data frame that contains the prevalence of COVID-19 for the Dutch population.
We wrap the data frame in a list with the name of the target population. NB: We could also include
lists with data frames for multiple populations. The name of the prevalence column must be `percent`.

```{r}
prevalence_df <- preprocessed_df |>
  filter(Measure == "Prevalence" & Metric == "Percent") |>
  select(age = age_num, sex = sex_num, percent = Value)

prevalence_list <- list(NL = list(data = prevalence_df))

```

Similarly, we create a data frame and list for the incidence. Note that the column name in the data
frame must be `value` instead of `percent`.

```{r}
incidence_df <- preprocessed_df |>
  filter(Measure == "Incidence" & Metric == "Percent") |>
  select(age = age_num, sex = sex_num, value = Value)

incidence_list <- list(NL = list(data = incidence_df))

```

We also create a data frame and for the disease excess mortality. Since people can be cured from
COVID-19, we use the `curedfraction` column to store the faction of the population that is cured 
from the disease. We estimate the cured fraction from the deaths but we could also obtain this data 
from other GBD data sources.
The column `acutelyfatal` is set to 0 for all rows because DYNAMO-HIA only allow either acutely 
fatal or cured fractions to be non-zero. We add two additional values to the list: `unit_type`
which indicates that we are using rates, and `parameter_type` which indicates that we supply the
cured fraction.

```{r}
mortality_df <- preprocessed_df |>
  mutate(acutelyfatal = 0,
         curedfraction = Value * 100000) |>
  filter(Measure == "Deaths" & Metric == "Rate") |>
  select(age = age_num, sex = sex_num, unit = Value, acutelyfatal, curedfraction) 

mortality_list <- list(NL = list(data = mortality_df,
                                 unit_type = "Rate",
                                 parameter_type = "Cured Fraction"))

```

Finally, we create a data frame and list for disease-adjusted life years (DALYs).

```{r}
daly_df <- preprocessed_df |>
  filter(Measure == "DALYs (Disability-Adjusted Life Years)" & Metric == "Percent") |>
  select(age = age_num, sex = sex_num, percent = Value) 

daly_list <- list(NL = list(data = daly_df))

```

With the prevalence, incidence, mortality, and DALY lists, we can now write the XML configuration 
files for the DYNAMO-HIA model that contain the reference data. 
We use the `write_disease_dir` function inside a temporary directory:

```{r}
withr::with_tempdir({
  fs::dir_create("Reference_Data")
  
  withr::with_dir("Reference_Data", {
    fs::dir_create("Diseases")
    
    withr::with_dir("Diseases", {
      write_disease_dir(disease_name = disease,
                        prevalences_list = prevalence_list,
                        incidences_list = incidence_list,
                        excess_mortalities_list = mortality_list, 
                        disability_list = daly_list,
                        risk_factor_list = list(),
                        diseases_list = list())
    })
  })
})

```

The function writes the XML files in the `Reference_Data/Diseases/` sub directory where the
DYNAMO-HIA Java application can find them. It also validates the XML schemata of the files so that 
they correspond to the expected format. Note that we did not include relative risks from risk
factors or other diseases (they go in the `risk_factor_list` and `diseases_list` arguments, 
respectively).
